Finding Subgraphs With Side Constraints\thanks{This research was supported by the Engineering and Physical Sciences Research Council [grant number EP/P026842/1]}

### 2021-01-28 paper/paper.tex
Final changes

\author{
    \"Ozg\"ur Akg\"un\inst{1}\orcidID{0000-0003-4460-6912} \and Jessica
    Enright\inst{2}\orcidID{0000-0002-0266-3292} \and Christopher
    Jefferson\inst{1}\orcidID{0000-0003-2979-5989} \and Ciaran
    McCreesh\inst{2}\orcidID{0000-0002-6106-4871} \and Patrick
    Prosser\inst{3}\orcidID{0000-0003-4460-6912} \and Steffen
    Zschaler\inst{4}\orcidID{0000-0001-9062-6637} \\
}
\authorrunning{\"O. Akgun et al.}
\institute{
    University of St Andrews, Scotland \and
    University of Glasgow, Scotland \\ \email{ciaran.mccreesh@glasgow.ac.uk} \and
    Algorithmicists Anonymous, Scotland \and
    King's College London, England
}

@@

We can express these problems in the Essence high level modelling language, as follows. We assume
vertices take their labels from the set $L = \{ 1\ldots\ell \}$ for some given $\ell$, and edges
from $E = \{ 1\ldots{}e \}$ (and so $\ell$ and / or $e$ may be 1, for applications that do not use
labels on vertices and / or edges):
\begin{lstlisting}
given l, e : int
letting L be domain int(1..l)
letting E be domain int(1..e)
\end{lstlisting}
We take as input a pattern directed graph which has $p$ vertices (which we number from 1 to $p$, in
the set $P$), and a target directed graph which has $t$ vertices (numbered from 1 to $t$, the set
$T$). Each graph is represented as total function from vertices to vertex labels, and a
\emph{partial} function from pairs of (not necessarily distinct) vertices to edge labels:
\begin{lstlisting}
given p, t : int
letting P be domain int(1..p)
letting T be domain int(1..t)

@@

We chose to use the Essence high-level constraint modelling language
\cite{DBLP:journals/constraints/FrischHJHM08} because of its support for convenient high-level types
like functions and relations. In a conventional modelling pipeline, problems are specified in
Essence and then are converted to concrete models that can be solved by a CP solver (in our
experiments Minion) via the Conjure and SavileRow tools. We augmented Conjure with a command line
option that instructs it to generate an extra file which describes how the variables representing
the graph are represented in the SavileRow input (known as Essence'). This is used by the graph
solver so it can map between its internal state and the variables in Essence'.  SavileRow converts
the graph model to Minion input, and this conversion contains information which allows Minion to map
its internal state back to the Essence' given to SavileRow. The graph solver and Minion then
communicate mappings between the nodes and edges in the graph using the identifiers in the Essence'
representation of the problem.

@@

The high-level approach, then, gives us a way of setting up the subgraph solver and a CP solver such
that they both have an equivalent same set of variables and values for the graph part of the model,
and tells us how to form a correspondence between their internal representations. Importantly, this
allows the CP solver to have additional variables that the subgraph solver does not know about, and
we do not specifically require the CP solver to be aware of all of the graph constraints.
(Additionally, due to preprocessing, the CP solver may also sometimes have only a subset of values
for some graph variables visible to it.)

@@

To enable communication between the two solvers, we use FIFOs (named pipes), and a simple text-based
protocol.  Both solvers are run and initialised, and then the subgraph solver proceeds as normal,
whilst the CP solver waits to be given commands. The subgraph solver then communicates its trail or
a candidate solution as a set of assignment constraints to the CP solver. The CP solver then sets
these assignments its state and then either performs a single propagation, or complete search, as
directed. When finished the solver communicates its success or failure state, and any deletions (if
requested), back to the subgraph solver and reverts any changes made by setting the assignments.
This approach is designed to be solver-agnostic, and adding support for different CP solvers (or
non-CP solvers) is simple, as long as they support performing a search or propagation from a given
set of assignments.

@@

\begin{figure}[p]
    \includegraphics{gen-graph-nosideconstraints.pdf}\hfill\includegraphics{gen-graph-oddeven.pdf}
    \bigskip
    \includegraphics{gen-graph-mostlyodd.pdf}\hfill\includegraphics{gen-graph-parity.pdf}
    \caption{Comparing different approaches to hybrid solving, showing the cumulative number of
    instances solved over time. On the top row, ``no side constraints'' then with the ``more odd
    target vertices than even target vertices'' side constraint; on the bottom row, the ``mostly odd
    target vertices'' side constraint on the left, and the ``odd to odd, even to even'' side
    constraint on the right. In the top left plot, the ``Checking'' and ``Rollback'' lines are
    indistinguishable.}\label{figure:cumulative}
\end{figure}

@@

As a point of comparison, we also implemented these three sets of side constraints natively inside
the Glasgow solver. For the ``more odd than even'' case, we did this through solution checking; for
``less than three odd'' we implemented checking during search; and for ``odd to odd, even to even''
we implemented initial domain filtering. Our implementation choices here are intended to reflect what a
reasonable programmer would do if the high level approach were not available (and we intentionally
selected side constraints that would not be too difficult to implement). In two of the three cases,
the hand-crafted code does somewhat outperform the hybrid solving, but in the ``more odd than even''
case, hybrid solving actually beats the hand-crafted dedicated solver implementation when using the
rollback approach, which we now describe.

@@

A less strict kind of subisomorphism is an \emph{offset}, where edge labels must match exactly but
offset by an integer constant. In our example, this means ``find a mapping where the event from 2
to 3 occurs one time unit after the event from 1 and 2''. We can model this as follows:
\begin{lstlisting}
find o : int(-e..e)
such that forAll (a,b) in defined(pat) .
    pat((a,b)) = o + tgt((f(a), f(b)))
\end{lstlisting}
and we find one additional solution,
\begin{lstlisting}
(1 --> 1, 2 --> 3, 3 --> 5)
\end{lstlisting}

>>>>

\author{
    \"Ozg\"ur Akg\"un\inst{1}\orcidID{0000-0003-4460-6912} \and Jessica
    Enright\inst{2}\orcidID{0000-0002-0266-3292} \and Christopher
    Jefferson\inst{1}\orcidID{0000-0003-2979-5989} \and Ciaran
    McCreesh\inst{2}\orcidID{0000-0002-6106-4871} \and Patrick
    Prosser\inst{3}\orcidID{0000-0003-4460-6912} \and Steffen
    Zschaler\inst{4}\orcidID{0000-0001-9062-6637} \\
}
\authorrunning{\"O. Akg\"un et al.}
\institute{
    University of St Andrews, Scotland \and
    University of Glasgow, Scotland \\ \email{ciaran.mccreesh@glasgow.ac.uk} \and
    Algorithmicists Anonymous, Scotland \and
    King's College London, England
}

@@

We can express these problems in the Essence modelling language, as follows. We assume
vertices take their labels from the set $L = \{ 1\ldots\ell \}$ for some given $\ell$, and edges
from $E = \{ 1\ldots{}e \}$ (and so $\ell$ and / or $e$ may be 1, for applications that do not use
labels on vertices and / or edges):
\begin{lstlisting}
given l, e : int
letting L be domain int(1..l)
letting E be domain int(1..e)
\end{lstlisting}
We take as input a directed pattern graph which has $p$ vertices (which we number from 1 to $p$, in
the set $P$), and a directed target graph which has $t$ vertices (numbered from 1 to $t$, the set
$T$). Each graph is represented as total function from vertices to vertex labels, and a
\emph{partial} function from pairs of (not necessarily distinct) vertices to edge labels:
\begin{lstlisting}
given p, t : int
letting P be domain int(1..p)
letting T be domain int(1..t)

@@

We chose to use the Essence modelling language \cite{DBLP:journals/constraints/FrischHJHM08} because
of its support for convenient high-level types like functions and relations, which can easily
describe graphs and related abstractions. In a conventional modelling pipeline, problems are
specified in Essence and then are converted to concrete models that can be solved by a CP solver (in
our experiments Minion) via the Conjure and SavileRow tools. We augmented Conjure with a command
line option that instructs it to generate an extra file which describes how the variables
representing the graph are represented in the SavileRow input (known as Essence'). This is used by
the graph solver so it can map between its internal state and the variables in Essence'.  SavileRow
converts the graph model to Minion input, and this conversion contains information which allows
Minion to map its internal state back to the Essence' given to SavileRow. The graph solver and
Minion then communicate mappings between the nodes and edges in the graph using the identifiers in
the Essence' representation of the problem.

@@

The high-level approach, then, gives us a way of setting up the subgraph solver and a CP solver such
that they both have an equivalent set of variables and values for the graph part of the model,
and tells us how to form a correspondence between their internal representations. Importantly, this
allows the CP solver to have additional variables that the subgraph solver does not know about, and
we do not specifically require the CP solver to be aware of all of the graph constraints.
(Additionally, due to preprocessing, the CP solver may also sometimes have only a subset of values
for some graph variables visible to it.)

@@

To enable communication between the two solvers, we use FIFOs (named pipes), and a simple text-based
protocol.  Both solvers are run and initialised, and then the subgraph solver proceeds as normal,
whilst the CP solver waits to be given commands. The subgraph solver then communicates its trail or
a candidate solution as a set of assignment constraints to the CP solver. The CP solver then sets
these assignments as its state and either performs a single propagation, or complete search, as
directed. When finished the solver communicates its success or failure state, and any deletions (if
requested), back to the subgraph solver and reverts any changes made by setting the assignments.
This approach is designed to be solver-agnostic, and adding support for different CP solvers (or
non-CP solvers) is simple, as long as they support performing a search or propagation from a given
set of assignments.

@@

\begin{figure}[p]
    \includegraphics{gen-graph-nosideconstraints.pdf}\hfill\includegraphics{gen-graph-oddeven.pdf}

    \bigskip

    \includegraphics{gen-graph-mostlyodd.pdf}\hfill\includegraphics{gen-graph-parity.pdf}
    \caption{Comparing different approaches to hybrid solving, showing the cumulative number of
    instances solved over time. On the top row, ``no side constraints'' then with the ``more odd
    target vertices than even target vertices'' side constraint; on the bottom row, the ``mostly odd
    target vertices'' side constraint on the left, and the ``odd to odd, even to even'' side
    constraint on the right. In the top left plot, the ``Checking'' and ``Rollback'' lines are
    indistinguishable. The Glasgow+ lines show the Glasgow Subgraph Solver with
    manually-implemented side constraints.}\label{figure:cumulative}
\end{figure}

@@

As a point of comparison, we also implemented these three sets of side constraints natively inside
the Glasgow solver: we label these as Glasgow+. For the ``more odd than even'' case, we did this
through solution checking; for ``less than three odd'' we implemented checking during search; and
for ``odd to odd, even to even'' we implemented initial domain filtering. Our implementation choices
here are intended to reflect what a reasonable programmer would do if the high level approach were
not available (and we intentionally selected side constraints that would not be too difficult to
implement). In two of the three cases, the hand-crafted code does somewhat outperform the hybrid
solving, but in the ``more odd than even'' case, hybrid solving actually beats the hand-crafted
dedicated solver implementation when using the rollback approach, which we now describe.

@@

A less strict kind of subisomorphism is an \emph{offset}, where edge labels must match exactly, but
offset by an integer constant. In our example, this means ``find a mapping where the event from 2
to 3 occurs one time unit after the event from 1 and 2''. We can model this as follows:
\begin{lstlisting}
find o : int(-e..e)
such that forAll (a,b) in defined(pat) .
    pat((a,b)) = o + tgt((f(a), f(b)))
\end{lstlisting}
and we find one additional solution,
\begin{lstlisting}
(1 --> 1, 2 --> 3, 3 --> 5)
\end{lstlisting}
